canaper: Categorical analysis of neo- and paleo-endemism in R

### 2023-04-07 ms/ms.Qmd
Update MS

To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
We used the `curveball` randomization algorithm with 50,000 iterations, since `cpr_iter_sim()` indicated this was sufficient to randomize the matrix (see "Major Functions").
Once the randomization algorithm and its settings have been selected, CANAPE can be conducted with just two commands.

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # Use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

@@

The `canaper` package enables CANAPE completely within R for the first time.
This will simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
<!-- FIXME: Add new studies -->
`canaper` has already been used at least three studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a].
We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.

@@

To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':

Nitta, J. H. et al. 2023. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD REFERENCE DETAILS HERE (version 1.0.0)

@@

**@fig-acacia-cpr.** Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007]; units in m.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type. This figure reproduces Figure 3 of @Mishler2014.

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

>>>>

To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v4.2.1.
We used the `curveball` randomization algorithm with 50,000 iterations, since `cpr_iter_sim()` indicated this was sufficient to randomize the matrix (see "Major Functions").
Once the randomization algorithm and its settings have been selected, CANAPE can be conducted with just two commands.

@@

# Fist do some data wrangling to make the results easier to plot
# (add lat/long columns)
acacia_canape <- acacia_canape |>
  separate(site, c("long", "lat"), sep = ":") |>
  mutate(across(c(long, lat), parse_number))
min_long <- min(acacia_canape$long)
min_lat <- min(acacia_canape$lat)

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # Use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  # add a 1000 km scalebar (data units are in m)
  geom_segment(
    x = min_long,
    y = min_lat,
    xend = min_long + 1000 * 1000,
    yend = min_lat
  ) +
  geom_text(
    x = min_long,
    y = min_lat,
    label = "1,000 km",
    vjust = -1, hjust = 0, size = 2
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

@@

The `canaper` package enables CANAPE completely within R for the first time.
This will simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
`canaper` has already been used at least five studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a; @Naranjo2023; @vanGalen2023].
We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.

@@

To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 0':

<!-- FIXME: update journal details when available -->
Nitta, J. H. et al. 2023. canaper: Categorical analysis of neo- and paleo-endemism in R. Ecography XX: XXX--XXX (version 0)

@@

**@fig-acacia-cpr.** Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007].
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type. This figure reproduces Figure 3 of @Mishler2014.

@@

min_long <- min(acacia_canape$long)
min_lat <- min(acacia_canape$lat)

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # Use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  # add a 1000 km scalebar
  geom_segment(
    x = min_long,
    y = min_lat,
    xend = min_long + 1000 * 1000,
    yend = min_lat
  ) +
  geom_text(
    x = min_long,
    y = min_lat,
    label = "1,000 km",
    vjust = -1, hjust = 0, size = 2
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

### 2023-04-07 ms/si.Qmd
Update MS

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  scale_fill_manual(values = cpr_endem_cols_4) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.background = element_rect(fill = "white", color = "white"),
    panel.grid.major = element_line(color = "grey60"),
    panel.grid.minor = element_blank()
  )

@@

**Fig. S2**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors (`cpr_endem_cols_4` palette in `canaper`).
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system; units in m.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.

>>>>

min_long <- min(acacia_canape$long)
min_lat <- min(acacia_canape$lat)

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  scale_fill_manual(values = cpr_endem_cols_4) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  # add a 1000 km scalebar (data units are in m)
  geom_segment(
    x = min_long,
    y = min_lat,
    xend = min_long + 1000 * 1000,
    yend = min_lat
  ) +
  geom_text(
    x = min_long,
    y = min_lat,
    label = "1,000 km",
    vjust = -1, hjust = 0, size = 2
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.background = element_rect(fill = "white", color = "white"),
    panel.grid.major = element_line(color = "grey60"),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

@@

**Fig. S2**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors (`cpr_endem_cols_4` palette in `canaper`).
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.
